Efficient implementation of penalized regression\\for genetic risk prediction

### 2018-08-29 paper/draft.tex
save before Biorxiv

\title{Efficient implementation of penalized regression for genetic risk prediction}
\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}

@@

\abstract{
Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs might have a major public health impact, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.  The ``Clumping+Thresholding'' (C+T) approach, which is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it easy to use and fast. However, previous work showed that jointly estimating SNP effects for computing PRS have the potential to significantly improve the predictive performance of PRS as compared to C+T.

@@

We compare the performance of penalized logistic regression to the C+T method and to a derivation of random forests. 
Penalized logistic regression consistently achieves higher predictive performance than the two other methods while being very fast. We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; AUC values increase from 83\% with the best prediction of C+T to 92.5\% with penalized logistic regression. We confirm these results in a data analysis of a case-control study for celiac disease where penalized logistic regression and the standard C+T method achieve AUC of 89\% and of 82\%.

In conclusion, our study demonstrates that penalized logistic regression is applicable to large-scale individual-level data and can achieve more discriminative polygenic risk scores. Our implementation is publicly available in the R package bigstatsr.\\

@@

Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs are useful for genetic epidemiology when testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}. Personalized medicine is another major application of PRSs. Personalized medicine envisions to use PRSs in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.

Several methods have been developed to predict disease status, or more generally any phenotype, based on SNP information. A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique uses GWAS summary statistics only, allowing for fast implementation. However, the ``C+T'' approach also has several limitations. Previous studies showed that predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}. Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRSs for complex human diseases by jointly estimating SNP effects. Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.

@@

To perform simulations, we use real genotype data and simulate new phenotypes \cite[]{zhou2013polygenic}. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability. We also compare different models for simulating phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
Overall, we found that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being very fast. This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.

@@

We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}. Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}. For simulations presented later, we first restrict this dataset to controls in order to remove the genetic structure induced by the celiac disease status. Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on a robust Mahalanobis distances computed using the first 10 principal components of the remaining individuals. This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing''). We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls). Both datasets contain 281,122 SNPs.

@@

In this study, we use two different measures of predictive accuracy. First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,lusted1971signal}. In the case of our study, the AUC is the probability that the PRS of a case is greater than the PRS of a control.
This measure indicates the extent to which we can distinguish between cases and controls using PRSs. As a second measure, the partial AUC for specificities between 90\% and 100\% is also reported \cite[]{dodd2003partial,mcclish1989analyzing}. This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings. 
When reporting AUC results of simulations, we use estimates of maximum achievable AUC of 84\% and 94\% for heritabilities of respectively 50\% and 80\%, based on three different yet consistent estimations (see Supplementary Materials). %Note that we also report the timing of the main computations and the number of SNPs used in the predictions.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.

For the C+T method, we report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``PRS-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``PRS-stringent''), and one that maximizes the AUC (``PRS-max'') for these two thresholds ($0$ and $5 \cdot 10^{-8}$) and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
As we report the optimal threshold based on the test set, the corresponding AUC value is an upper bound of the AUC for the C+T method.

@@

Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use elastic-net regularization with $\alpha \in \{1, 0.5, 0.05, 0.001\}$ (determined thanks to an embedded grid search).

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr.
This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We refer to this method as ``logit-simple'' in the results section.

@@

When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}). 
Moreover, penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).

Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region. In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 83\% with ``PRS-max'', while the maximum achievable AUC is 94\% (Figure \ref{fig:main-AUC-logit}).

@@

Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the commonly used value of $0.2$. Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases tested in this paper (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})

Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).

@@

We tested the T-Trees method in scenario \textnumero1.
As compared to ``logit-simple'', T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}). Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using ``logit-simple'' (Figure \ref{fig:supp-ttrees}).

We also compared the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features (variables) coding for recessive and dominant effects.
Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}). For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', except when there are 3000 causal SNPs.
Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``logit-simple''.

@@

Joint logistic regressions provide higher AUC values for the Celiac data: 88.7\% with ``logit-simple'' and 89.1\% with ``logit-triple'' as compared to 82.5\% with the C+T method.
The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
Note that for the C+T method, we still report the best result among 102 p-value thresholds.
In terms of computation time, we report only the GWAS computation for the C+T method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once and testing different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (296 vs 190 seconds).

@@

In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently. Penalized logistic regression nearly always outperform the C+T method, and the benefits of using it are stronger with an increasing sample size or when causal SNPs are correlated with one another.

@@

The choice of hyper-parameter values is very important as it can greatly impact method performance. In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively. 
The choice of the $r^2$ threshold of the clumping step is important. Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in LD; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
Based on the simulations, we find that using stringent threshold ($r^2 = 0.05$) leads to better predictive performance, even when causal SNPs are correlated. It means that accurately estimating SNP effects is more important than including all causal SNPs.
Moreover, in this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only. The choice of this threshold is very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}.

On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model. Not only this procedure provides near-optimal results, but it also accelerates the training of the model thanks to the development of an early stopping criterion. 
Usually, cross-validation is used to choose hyper-parameter value and then the model is trained again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}. Yet, performing cross-validation and retraining is computationally demanding, CMSA offers a less burdensome alternative. 
Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
%Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model. Yet, if using e.g. $\alpha = 0.001$, it would lead to a similar model.

@@

Yet, this approach is not able to detect interaction-type effects. 
In order to capture interaction-type effects, we tested T-Trees, a method that is able to exploit SNP correlations thanks to special decision trees \cite[]{botta2014exploiting}. 
However, predictive performance of T-Trees were consistently lower than penalized logistic regression, even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle.

@@

Our approach has one major limitation: the main advantage of the C+T method is that it is applicable directly to summary statistics, allowing to leverage the largest GWAS sample size to date, even when individual cohort data cannot be merged because of practical and ethical reasons (e.g. consortium data including hundreds of cohorts). As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work. The current version is of particular interest for the analysis of modern SNP dataset including hundreds of thousands of individuals.

Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}. In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population. 
%We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}. Finally, we also wish to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.

>>>>

\title{Efficient implementation of penalized regression\\for genetic risk prediction}
\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}

@@

\abstract{
Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS might have a major public health impact, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.  
The ``Clumping+Thresholding'' (C+T) approach, which is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it fast and easy to use. 
However, previous work showed that jointly estimating SNP effects for computing PRS has the potential to significantly improve the predictive performance of PRS as compared to C+T.

@@

We compare the performance of penalized logistic regression to the C+T method and to a derivation of random forests. 
Penalized logistic regression consistently achieves higher predictive performance than the two other methods while being very fast. We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; AUC values increase from 83\% with the best prediction of C+T to 92.5\% with penalized logistic regression. We confirm these results in a data analysis of a case-control study for celiac disease where penalized logistic regression and the standard C+T method achieve AUC of 89\% and of 82.5\%.

In conclusion, our study demonstrates that penalized logistic regression is applicable to large-scale individual-level data and can achieve more discriminative polygenic risk scores. Our implementation is publicly available in R package bigstatsr.\\

@@

Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS are useful for genetic epidemiology when testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}. Personalized medicine is another major application of PRS. Personalized medicine envisions to use PRS in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
Yet, PRS would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.

Several methods have been developed to predict disease status, or more generally any phenotype, based on SNP information. A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique uses GWAS summary statistics only, allowing for fast implementation. However, the ``C+T'' approach also has several limitations. 
Previous studies have shown that predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}. Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects. Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.

@@

To perform simulations, we use real genotype data and simulate new phenotypes \cite[]{zhou2013polygenic}. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability. We also compare different models for simulating phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
Overall, we find that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being very fast. This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.

@@

We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}. Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}. For simulations presented later, we first restrict this dataset to controls in order to remove the genetic structure induced by the celiac disease status. Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on a robust Mahalanobis distances computed using the first 10 principal components of the remaining individuals. This 3-step filtering process results in a sample of 7100 individuals with minimal population structure (see supplementary notebook ``preprocessing''). We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls). Both datasets contain 281,122 SNPs.

@@

In this study, we use two different measures of predictive accuracy. First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,lusted1971signal}. In the case of our study, the AUC is the probability that the PRS of a case is greater than the PRS of a control.
This measure indicates the extent to which we can distinguish between cases and controls using PRS. 
As a second measure, we also report the partial AUC for specificities between 90\% and 100\% \cite[]{dodd2003partial,mcclish1989analyzing}. This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings. 
When reporting AUC results of simulations, we use estimates of maximum achievable AUC of 84\% and 94\% for heritabilities of respectively 50\% and 80\%. These estimates are based on three different yet consistent estimations (see Supplementary Materials). %Note that we also report the timing of the main computations and the number of SNPs used in the predictions.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$. We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``PRS-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``PRS-stringent''), and one that maximizes the AUC (``PRS-max'') for these two thresholds ($0$ and $5 \cdot 10^{-8}$) and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
As we report the optimal threshold based on the test set, the AUC for ``PRS-max'' is an upper bound of the AUC for the C+T method.

@@

Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use an embedded grid search over $\alpha \in \{1, 0.5, 0.05, 0.001\}$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of L1-regularization, the resulting vectors of coefficients are sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We refer to this method as ``logit-simple'' in the results section.

@@

When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-PRS}). 
Moreover, penalized logistic regression consistently provides higher predictive performance than the C+T method across all scenarios we considered, excepted in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).

Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region. In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 84\% with ``PRS-max'', while the maximum achievable AUC is 94\% (Figure \ref{fig:main-AUC-logit}).

@@

Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the common value of $0.2$. Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases tested in this paper (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})

Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the implementation of R package biglasso \cite[]{zeng2017efficient}, only slightly better for biglasso, which is likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).

@@

We tested the T-Trees method in scenario \textnumero1.
As compared to ``logit-simple'', T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}). 
Even when simulating a ``fancy'' model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using ``logit-simple'' (Figure \ref{fig:supp-ttrees}).

We also compared the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features (variables) coding for recessive and dominant effects.
Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}). For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', excepted when there are 3000 causal SNPs.
Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``simple'' solution.

@@

Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with ``logit-simple'' and 89.1\% with ``logit-triple'' as compared to 82.5\% with the C+T method.
The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
Note that for the C+T method, we still report the best result among 102 p-value thresholds.
In terms of computation time, we report only the GWAS computation for the C+T method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once and testing different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (296 vs 190 seconds).

@@

In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently. 
Penalized logistic regression nearly always outperform the C+T method, and the benefits of using it are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.

@@

The choice of hyper-parameter values is very important since it can greatly impact method performance. In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively. 
The choice of the $r^2$ threshold of the clumping step is important. Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in Linkage Desiquilibrium; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated. It means that accurately estimating SNP effects is more important than including all causal SNPs.
Moreover, in this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only. The choice of this threshold is very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}.

On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model. Not only this procedure provides near-optimal results (as compared to the best prediction when using R package biglasso), but it also accelerates the training of the model thanks to the development of an early stopping criterion. 
Usually, cross-validation is used to choose hyper-parameter values and then the model is trained again with these particular hyper-parameter values \cite[]{Hastie2008,wei2013large}. Yet, performing cross-validation and retraining the model is computationally demanding; CMSA offers a less burdensome alternative. 
Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
%Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model. Yet, if using e.g. $\alpha = 0.001$, it would lead to a similar model.

@@

Yet, this approach is not able to detect interaction-type effects. 
In order to capture interaction-type effects, we tested T-Trees, a method that is able to exploit SNP correlations thanks to special decision trees \cite[]{botta2014exploiting}. 
However, predictive performance of T-Trees were consistently lower than with penalized logistic regression, even when simulating a model with dominant and interaction-type effects that T-Trees should be able to handle.

@@

Our approach has one major limitation: the main advantage of the C+T method is that it is applicable directly to summary statistics, allowing to leverage the largest GWAS sample size to date, even when individual cohort data cannot be merged because of practical and ethical reasons (e.g. consortium data including many cohorts). As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work. The current version is of particular interest for the analysis of modern SNP dataset including hundreds of thousands of individuals.

Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}. In next study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population. 
%We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}. Finally, we also wish to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
